---
title: "Exploratory Data Analysis for Global Mean Sea Level Data"
author: "Thad Hoskins, Gina McFarland, Paul O'Leary"
output:
  word_document: default
  html_document:
  df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(ggpubr)
library(smooth)
library(forecast)
```

Following is an analysis of variables to help determine which would be the most appropriate response variable to analyze

```{r}
# V2 vs V6  : merged file cycle number vs. GMSL (Global Isostatic Adjustment (GIA) not 
# applied) variation (mm) with respect to 20-year TOPEX/Jason collinear mean reference
# vector of the widths of the fields
wths <- c(3,5,14,10,10,10,10,10,10,10,10,10)

datProj <- read.fwf("GMSL_DATA.txt", wths, header = FALSE)

gPv6 <- ggplot(datProj, aes(x=V2, y=V6))
gPv6 <- gPv6 + geom_point(size = .3)
gPv6
```
Merged File Cycle # (V2) vs. Raw Data (Column V6) 

```{r}
# V6    : year+fraction of year (mid-cycle) vs.  GMSL (Global Isostatic Adjustment (GIA) not 
# applied) variation (mm) with respect to 20-year TOPEX/Jason collinear mean reference
# vector of the widths of the fields
wths <- c(3,5,14,10,10,10,10,10,10,10,10,10)

datProj <- read.fwf("GMSL_DATA.txt", wths, header = FALSE)

gPv6 <- ggplot(datProj, aes(x=V3, y=V6))
gPv6 <- gPv6 + geom_point(size = .3)
gPv6
```
Time (V3) vs. Raw Data (Column V6)


```{r}
#V7 merged file cycle # vs. standard deviation of GMSL (GIA not applied) variation estimate (mm)
# vector of the widths of the fields
wths <- c(3,5,14,10,10,10,10,10,10,10,10,10)

datProj <- read.fwf("GMSL_DATA.txt", wths, header = FALSE)

gPv7 <- ggplot(datProj, aes(x=V3, y=V7))
gPv7 <- gPv7 + geom_point(size = .3)
gPv7
```
```{r}
#V7: time  vs. standard deviation of GMSL (GIA not applied) 
# variation estimate (mm)
# vector of the widths of the fields
wths <- c(3,5,14,10,10,10,10,10,10,10,10,10)

datProj <- read.fwf("GMSL_DATA.txt", wths, header = FALSE)

gPv7 <- ggplot(datProj, aes(x=V3, y=V7))
gPv7 <- gPv7 + geom_point(size = .3)
gPv7
```
Time (V3) vs. Standard Deviation of GMSL (Column V7)

```{r}
#V8: Time # vs. 8 smoothed (60-day Gaussian type filter) GMSL (GIA not applied) variation (mm)  with respect
#to 20-year mean

# vector of the widths of the fields
wths <- c(3,5,14,10,10,10,10,10,10,10,10,10)

datProj <- read.fwf("GMSL_DATA.txt", wths, header = FALSE)

gPv8 <- ggplot(datProj, aes(x=V3, y=V8))
gPv8 <- gPv8 + geom_point(size = .3)
gPv8
```
Time (V3) vs.Smoothed (60-day Gaussian type filter) GMSL (Column V8)

```{r}
#V9: Time vs. GMSL (Global Isostatic Adjustment (GIA) applied) variation (mm) )  with respect to 20-year mean

# vector of the widths of the fields
wths <- c(3,5,14,10,10,10,10,10,10,10,10,10)

datProj <- read.fwf("GMSL_DATA.txt", wths, header = FALSE)

gPv9 <- ggplot(datProj, aes(x=V3, y=V9))
gPv9 <- gPv9 + geom_point(size = .3)
gPv9
```

Time (V3) vs. GMSL (Global Isostatic Adjustment (GIA) applied) variation (mm)) with respect to 20-year mean (V9)


```{r}
#V10: Time (V3) vs. standard deviation of GMSL (GIA applied) variation estimate (mm)

# vector of the widths of the fields
wths <- c(3,5,14,10,10,10,10,10,10,10,10,10)

datProj <- read.fwf("GMSL_DATA.txt", wths, header = FALSE)

gPv10 <- ggplot(datProj, aes(x=V3, y=V10))
gPv10 <- gPv10 + geom_point(size = .3)
gPv10
```

 Time (V3) vs. standard deviation of GMSL (GIA applied) variation estimate (mm) (v10)

```{r}
#V11: Time (V3) vs. smoothed (60-day Gaussian type filter) GMSL (GIA applied) variation (mm)) with respect to 20-year mean

# vector of the widths of the fields
wths <- c(3,5,14,10,10,10,10,10,10,10,10,10)

datProj <- read.fwf("GMSL_DATA.txt", wths, header = FALSE)

gPv11 <- ggplot(datProj, aes(x=V3, y=V11))
gPv11 <- gPv11 + geom_point(size = .3)
gPv11
```
Time (V3) vs. smoothed (60-day Gaussian type filter) GMSL (GIA applied) variation (mm)) with respect to 20-year mean (V11)


```{r}
#V12: Time vs. 12 smoothed (60-day Gaussian type filter) GMSL (GIA applied) variation (mm); annual and semi-annual signal removed )  with respect to 20-year mean
# vector of the widths of the fields
wths <- c(3,5,14,10,10,10,10,10,10,10,10,10)

datProj <- read.fwf("GMSL_DATA.txt", wths, header = FALSE)

gPv12 <- ggplot(datProj, aes(x=V3, y=V12))
gPv12 <- gPv12 + geom_point(size = .3)
gPv12
```
Time (V3) vs. 12 smoothed (60-day Gaussian type filter) GMSL (GIA applied) variation (mm); annual and semi-annual signal removed) (V12) 

Field 6 has the raw data, and fields 7-12 have varying analyses on the raw data. Unsurprisingly, all
graphs show similar trends. We concluded that it was best to study the raw data as opposed to data that
has had any sort of manipulation or filtering for this analysis.

```{r}
length(datProj$V6)
```
1006 observations in the dataset
```{r}
summary(datProj$V6)
```
#GMSL (Global Isostatic Adjustment (GIA) not applied) variation (mm) with respect to 20-year TOPEX/Jason collinear mean reference
#The TOPEX/Jason collinear mean reference is zero and these measurements are +/- from that point in mm

There is data below 0. This makes sense because the TOPEX/Jason collinear mean reference is zero and these measurements are +/- from that point in mm.
```{r}
head(datProj$V6)
tail(datProj$V6)
```

```{r}
hist(datProj$V6)
```

```{r}
ggqqplot(datProj$V6)
```
While the data in the middle looks to be normally distributed, the ends look problematic.

```{r}
# Missing data code: 99900.000
sum(datProj$V6 == 99900.000)
sum(datProj$V3 == 99900.000)

sum(is.na(datProj$V6))
sum(is.na(datProj$V3))

```
No missing data or NAs in our variables of interest


```{r}
outliers <- boxplot(datProj$V6)$out
outliers
```
No outliers in our raw data

```{r}

g2 <- ggplot() + geom_line(data = datProj, aes(x=V3, y=V6)) + 
      geom_line(aes(x = datProj$V3,y = ma(datProj$V6,order = 18)), color = "blue")
g2

```
Smoothed to a moving average of 2 measurements per year. 